Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-19 submission/manuscript.Rmd
small edits

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms. Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data indicate that KDM5C decommissions germline genes by facilitating CGI methylation to prevent runaway activation of germline developmental programs in somatic lineages.

@@

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated germline-enriched gene list. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes.

@@

Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D). Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of gonad-enriched genes is independent of organismal sex.

Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E). Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.

@@

_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C). We then identified tissue-enriched DEGs in a RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D). Using our list of mouse germline-enriched genes assembled above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.

@@

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H). These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate and later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell state. Altogether, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes, indicating germline gene misexpression mirrors germline development during the progression of somatic development.

@@

The majority of germline genes expressed in both sexes were more highly dysregulated in females compared to males (Figure 4D-F). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression. Female sensitivity could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G). The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of XCI defects.

@@

To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1). KDM5C was primarily localized to gene promoters in both cell types (Promoters = transcription start site (TSS) &#177; `r promorange`bp, EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)`), although PNCs showed increased localization to non-promoter regions (Figure 5A).

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were bound by KDM5C only in EpiLCs (3,677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for PNC-specific KDM5C target genes. Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E). Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C). Representative examples of KDM5C-bound and unbound EpiLC DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary figure 1A). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their gene promoters, but genes dysregulated across _Kdm5c_-KO development are often direct KDM5C targets.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression of distinct germline gene classes within somatic lineages. Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. While sex did not determine whether egg or sperm-specific genes were dysregulated, it greatly influenced the degree of germline gene dysregulation. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous knockout females also had over double the number of germline DEGs than hemizygous knockout males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males. Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D. However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

>>>>

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms. Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) that is mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene repression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a break against runaway activation of germline developmental programs in somatic lineages.

@@

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for germ cell development, given the lack of a comprehensive list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic versus spermatid differentiation genes.

@@

Interestingly, we also observed significant enrichment of ovary-biased genes in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D). Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of gonad-enriched genes is independent of organismal sex.

Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

Analysis of oligo(dT)-primed libraries[@vallianatosMutuallySuppressiveRoles2020] demonstrates aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts in the _Kdm5c_-KO brain (Figure 1C-E). Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.

@@

_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C). We then identified tissue-enriched DEGs in a RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D). Using our list of mouse germline-enriched genes assembled above, we identified `r nrow(epigerm)` germline genes misexpressed in male _Kdm5c_-KO EpiLCs.

@@

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H). These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate and again later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell state. Altogether, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes, indicating germline gene misexpression mirrors germline development during the progression of somatic development.

@@

The majority of germline genes expressed in both sexes were more highly dysregulated in females compared to males (Figure 4D-F). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression. Heightened germline gene dysregulation in females could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, female germline DEGs were not biased towards the X chromosome and females had a a similar overall proportion of germline DEGs belonging to the X chromosome as males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G). The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of XCI defects.

@@

To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016] (MACS2 q < 0.1, fold enrichment > 1, and removal of false-positive peaks in _Kdm5c_-KO). EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`). KDM5C was primarily localized to gene promoters in both cell types (promoters = transcription start site (TSS) &#177; `r promorange` bp, EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)`), although PNCs showed increased localization to non-promoter regions (Figure 5A).

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were bound by KDM5C only in EpiLCs (3,677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline ontologies were enriched only in EpiLC-specific, KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no significant ontologies for PNC-specific KDM5C target genes. Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E). Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C). Representative examples of EpiLC DEGs bound and unbound by KDM5C in EpiLCs are _Dazl_ and _Stra8_, repsectively (Figure 5F). However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary figure 1A). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their gene promoters, however genes dysregulated across _Kdm5c_-KO development are often direct KDM5C targets.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs have heightened germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression of distinct germline gene classes within somatic lineages. Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. While sex did not determine whether egg or sperm-specific genes were dysregulated, it greatly influenced the degree of germline gene dysregulation. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) . Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D. However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression. Altogether, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/the inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_. These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus. Thus, ectopic germline gene expression occurs in conjunction with overall proper somatic differentiation in _Kdm5c_-KO animals.
